File-based localization of numerical perturbations in data analysis pipelines

### 2020-03-31 paper.tex
Edits in results

Provenance graphs are often subject-dependent, due to variations in input data among subjects.
%, which justifies the processing
% of multiple subjects. For instance, data acquisitions may be
% repeated a different number of times in each subject, for noise-reduction
% purposes. 
Some of these differences can be neglected, for instance when a
data decompression step is present at the beginning of the execution for
some subjects only, but other differences cannot, for instance
when complete different processing paths are used in different subjects. \tristan{We need to conclude on that comment}

@@

We applied \toolname to find the origin of between-OS differences that
occur in many neuroimaging pipelines. In particular, we studied the minimal
pre-processing pipelines released by the Human Connectome Project
(\href{https://www.humanconnectome.org}{HCP}), a leading initiative in
neurosciences.

@@

The HCP developed a set of pre-processing pipelines to process structural,
functional, and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, and particularly
PreFreeSurfer and FreeSurfer. 
A detailed description of the analyses done in these
pipelines is available in~\cite{glasser2013}. 
In summary, PreFreeSurfer consists of the following steps: 
\begin{itemize}
\item Distortion Correction (DC), 
\item Anatomical Average (AAve), 
\item ACPC (Anterior Commissure, Posterior Commissure) Alignment (ACPC-A), 
\item Brain Extraction (BExt), 
\item Bias Field Correction (BFC), 
\item Atlas-Registration (AR).
\end{itemize}
And FreeSurfer consists of the following ones:
\begin{itemize}
\item Downsampling images, 
\item T1w image registration, 
\item Placement of surface images, 
\item Folding-based surface registration.
\end{itemize}
The average processing time per subject is approximately 2 hours for
PreFreeSurfer and 8 hours for FreeSurfer. The average output file size is
approximately 2.7~GB for PreFreeSurfer and 4.1~GB for FreeSurfer.
\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}

We randomly selected 20 unprocessed subjects 
from the HCP data release S500 
available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
\tristan{Can you let me know where the data is located, so that I could describe it here?
Subjects are located in consider:/data/asalari/ali-tests/PFS\_Centos6\_Traced
and consider:/data/asalari/ali-tests/PFS\_Centos7\_Traced 
and outputs are in consider:/data/asalari/ali-tests/nurm-out}
See~\cite{van2013wu} for a discussion of these various modalities.

@@

We built Docker images for the HCP pre-processing pipelines v3.19.0
(PreFreeSurfer and FreeSurfer) in CentOS 6.9 (Final) and CentOS 7.4 (Core),
a widely-used operating system in neurosciences
~\cite{hanke2011neuroscience}. Container images are available on
\href{https://hub.docker.com/r/bigdatalabteam/hcp-prefreesurfer/}{DockerHub}.
They contain the software dependencies of the HCP pipelines, including FSL
(version 5.0.6), FreeSurfer (version 5.3.0-HCP, CentOS4 build), and
Connectome Workbench (version 1.0).

We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2
CentOS versions. Each subject was processed twice on the same operating
system to detect potential within-OS variability, for instance coming from
pseudo-random operations.

\tristan{See what to do with this paragraph}
Within-OS variability may also be coming from other factors than
pseudo-randomness. For instance, the same files may have different metadata
such as timestamps, file paths, and other execution-specific information.
To ensure that the differences are not related to the metadata, we used
specific functions from FreeSurfer tool including \emph{mri\_diff},
\emph{mris\_diff}, and \emph{lta\_diff}. These functions compare files
based on the image dimensions, voxel intensities, the number of vertices,
and geometry information (see~\cite{fischl2012freesurfer} for more
details) \tristan{Ali, are these functions used by \toolname to compare results?
Yes, they have been used based on the file format}.

In addition, we used the Dice coefficient metric to evaluate the
similarity of image segmentation results, defined by:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\].

@@

Figure~\ref{fig:pfs_freq} shows the frequency of non-reproducible pipeline processes
by PreFreeSurfer pipeline step.
Differences were observed in linear registration 
with FSL FLIRT (in ACPC-Alignment, Brain Extraction, Distortion Correction, 
Atlas Registration), in non-linear registration with FSL FNIRT (in Brain Extraction 
and Altas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL maths  (in Anatomical Average). \tristan{figure and caption to be fixed.}

@@

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{images/pfs_heatmap.png}
  \caption{Frequency heatmap of non-reproducible PreFreeSurfer processes (N=20). 
           Rows refer to PreFreeSurfer pipeline steps.
           Cell labels describe the frequency distribution of the instances as ``N [M] x Y" 
           in which N is the proportion of the process 
           that creates differences out of all M times that is occurred. 
           Also, times Y shows the number of processes with the same distribution.
           For example, the FNIRT process occurred at most 6 times in sub-pipeline DC in each subject. 
           Also, this shows 3 different distributions among the subjects.
           The left cell shows 2 instances of the FNIRT that introduce differences in 19 out of 20 subjects.
           The middle cell indicates three instances of the FNIRT that create no differences for all 20 subjects.
           The right cell shows one instance of the FNIRT that introduces differences in 17 out of 20 subjects.
           The range between red and green colors shows the highest and lowest 
           proportion of occurrences of the processes with differences.
           Also, grey cells indicate that the tool is not executed in the sub-pipeline.}
  \label{fig:pfs_freq}
\end{figure*}

To give an idea of the observed differences, Figure~\ref{fig:fnirt_result} 
compares results of Brain Extraction in PreFreeSurfer. \tristan{Ali, there is no
FNIRT process in ACPC-Alignment, please check. Ali: yes, it was brain extraction.}

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/t2w_alignment.png} 
    \caption{Differences between FNIRT results from PreFreeSurfer 
    (checkerboard image \emph{T2w\_acpc\_to\_MNI\_nonli.nii.gz}) (CentOS6 vs. 
    CentOS7). An animated version of the comparison is available 
    \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/images/pfs_t2w_alignment.gif}
    {here}.
} 
    \label{fig:fnirt_result}
\end{figure}

@@

>>>>

Provenance graphs are often subject-dependent, due to variations in input data among subjects.
%, which justifies the processing
% of multiple subjects. For instance, data acquisitions may be
% repeated a different number of times in each subject, for noise-reduction
% purposes. 
Some of these differences can be neglected, for instance when a
data decompression step is present at the beginning of the execution for
some subjects only, but other differences cannot, for instance
when complete different processing paths are used in different subjects. \toolname 
includes helpers to identify different instances of provenance graphs \tristan{mention tree distance here}.

@@

We applied the \toolname to find the origin of between-OS differences that
frequently occur in neuroimaging pipelines. In particular, we studied the minimal
pre-processing pipelines released by the Human Connectome Project
(\href{https://www.humanconnectome.org}{HCP}), a leading initiative in
neurosciences.

@@

The HCP developed a set of pre-processing pipelines to process structural,
functional, and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, and particularly
on PreFreeSurfer and FreeSurfer. 
A detailed description of the analyses done by these
pipelines is available in~\cite{glasser2013}. 
In summary, PreFreeSurfer consists of the following steps: 
\begin{itemize}
\item Distortion Correction (DC), 
\item Anatomical Average (AAve), 
\item Anterior/Posterior Commissure Alignment (ACPC-A), 
\item Brain Extraction (BExt), 
\item Bias Field Correction (BFC), 
\item Atlas-Registration (AR).
\end{itemize}
And FreeSurfer consists of the following ones: \tristan{Check these steps}
\begin{itemize}
\item Image downsampling, 
\item T1w image registration, 
\item Surface placement, 
\item Surface registration.
\end{itemize}
The average processing time per subject is approximately 2~hours for
PreFreeSurfer and 8~hours for FreeSurfer. The average output file size is
approximately 2.7~GB for PreFreeSurfer and 4.1~GB for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}

We randomly selected 20 unprocessed subjects 
from the HCP data release S500 
available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
For each subject, available data consisted of 1 or 2 T1 images, and 1 or 2 T2 images. 
% \tristan{Can you let me know where the data is located, so that I could describe it here?
% Subjects are located in consider:/data/asalari/ali-tests/PFS\_Centos6\_Traced
% and consider:/data/asalari/ali-tests/PFS\_Centos7\_Traced 
% and outputs are in consider:/data/asalari/ali-tests/nurm-out}
Acquisition protocols and parameters are detailed in~\cite{van2013wu}. \tristan{Add summary of matrix sizes.}

@@

We built Docker images for the HCP pre-processing pipelines v3.19.0
(PreFreeSurfer and FreeSurfer) in CentOS 6.9 (Final) and CentOS 7.4 (Core), available on
\href{https://hub.docker.com/r/bigdatalabteam/hcp-prefreesurfer/}{DockerHub}.
Container images contain the HCP software dependencies, including FSL
(version 5.0.6), FreeSurfer (version 5.3.0-HCP, CentOS4 build), and
Connectome Workbench (version 1.0).

We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2
CentOS versions. Each subject was processed twice on the same operating
system to detect within-OS variability coming from pseudo-random
operations. We ignored execution-specific information such as file path or
timestamps by using FreeSurfer tools \texttt{mri\_diff},
\texttt{mris\_diff}, and \texttt{lta\_diff} that determine if MRI volumes
or transformations differ. To compare segmentations, we used the Dice coefficient defined as follows:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\].

@@

Figure~\ref{fig:pfs_freq} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
Differences were observed in linear registration 
with FSL FLIRT (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL FNIRT (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL maths  (in Anatomical Average). \tristan{figure and caption to be fixed.}

@@

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{images/pfs_heatmap.png}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps (N=20). 
  \tristan{Remove frequency values as they are shown by the color already. Remove multipliers, multiply cells instead. Cells should just contain subject numbers.}
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  number of subjects where the corresponding process appears. For example,
  the FLIRT tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 95\% of the subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  85\% of the subjects.
        }
  \label{fig:pfs_freq}
\end{figure*}

To visually illustrate the observed differences,
Figure~\ref{fig:fnirt_result} compares FNIRT results in Brain Extraction
for a particular subject. Differences appear to be very substantial, in
particular in the frontal and parietal lobes. 
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/t2w_alignment.png} 
    \caption{Differences between FNIRT results in PreFreeSurfer's Brain Extraction (CentOS6 vs. 
    CentOS7) \tristan{Improve contrast}. An animated version of the comparison is available 
    \href{https://github.com/ali4006/HCP-reproducibility-paper/blob/master/images/pfs_t2w_alignment.gif}
    {here}.
} 
    \label{fig:fnirt_result}
\end{figure}

@@

The reproducibility of a given tool may vary across subjects, and across parameters. 
See FLIRT in DC in Figure 4.
